Digit Recognition: Nearest Neighbors vs. Perceptron

### 2013-06-10 Paper/paper.tex
Done!

Optical Character Recognition (OCR) is a complex computer science problem which
has several effective machine learning solutions. We implemented the K Nearest
Neighbors (KNN) and Kernelized Perceptron algorithms and compared the
effectiveness of both methods in classifying a testing data set of 28,000
images. Each image is formatted as an array of 768 pixel values, with each value
indicating how dark each pixel was in the image. Both of our algorithms were
trained on a training data set of 42,000 images. We used cross validation to
tune the K value in KNN and to choose a kernel function and dimension value for
Perceptron. Despite its relative simplicity, we found that KNN was more
effective than Perceptron in classifying the digits. The nearest neighbors
algorithm correctly classified 96.857\% of the testing samples on a weighted
neighbors algorithm which looked at 4 neighbors. Our best Perceptron
implementation correctly classified 95.67\% of the testing samples using an
exponential kernel with a $\sigma = 5$. Despite its higher accuracy, the runtime
of KNN was 8x slower than that of the Perceptron.

@@

The Kaggle data provides two groups of data, 42,000 training samples with labels
and 28,000 unlabeled testing samples. Each training sample contains a 48 pixel
by 48 pixel image containing a handwritten number. The image is represented as
an array of 768 pixel values which represent how dark each pixel of the image
is. Table 1 is an example of a training sample.

@@

K Nearest Neighbors is a relatively simple algorithm which often gets surprisingly
good results given its simplicity. Given a testing sample $X$, the algorithm
looks through all the training samples it has previously seen and finds the
\textit{K} most similar samples. The algorithm then implements some form of
voting mechanism among the \textit{K} samples to classify the testing sample as
a digit 0-9. The voting mechanism can either be as straightforward as selecting the
majority label, to something more complicated which involves weighting the
samples based on some distance metric.

@@

@@

\begin{table}[ht]
\textbf{\caption{KNN Cross Validation Error Rates}} % title
\centering
\begin{tabular}{c | c c }
K & Unweighted KNN Error & Weighted KNN Error \\
\hline
1 & .077 & .077 \\
2 & .092 & .092 \\
3 & .075 & .075 \\
4 & .078 & .073 \\
5 & .078 & .075 \\
10 & .085 & .079 \\
15 & .095 & .089 \\
20 & .102 & .098 \\ [1ex]
\end{tabular}
\label{table:nonlin}
\end{table}

@@

Our cross validation results suggest that running weighted KNN with $K = 4$
yields the most successful classification rate. One caveat to this is that our
cross validation code ran using 5,000 training samples instead of the 42,000
samples in our full training set. It is possible that data we used to cross
validate is no representative of the full data set.

@@

Our KNN implementation was extremely successful, correctly classifying 96.857\%
of the testing samples. One downside to it however is the runtime. Training is
near instantaneous since the algorithm simply loads the data into an array. To
classify a sample the algorithm needs to compute the euclidean distance between
the testing sample and every training sample. In the case of the full data set
this is 42,000 samples, each of which has 768 dimensions. As a result, each
testing sample takes approximately. Altogether it takes close to 8 hours to
classify all 28,000 testing samples.

@@

One problem that needs to be addressed is that a standard perceptron is a binary
classifier meaning it selects between one of two classes. We resolved this issue
by training $10 \choose 2 = 45$ classifiers. Each classifier is responsible for
classifying between two digits. For example, we would train a perceptron to
classify between 0 and 1, and a second to classify between 0 and 2, and so
on. When we classify we vote between all of our trained classifiers and output
the majority element.

@@

The kernel function is represented by $K(X_i, X)$. There are a number of choices
for this kernel function.

@@

Finally, the runtime of the perceptron algorithm is a little over an hour. It
takes approximately 45 minutes to train and about 20 minutes to test the
data. The faster runtime allowed us to cross validate on the full data set.

@@

\begin{center}
\begin{tabular}{ c| c c}
	\hline
Algorithm & Percent Success & runtime\\
	\hline
Nearest Neighbors, $k = 4$   & 96.857\% & $\sim$8 hours\\
Perceptron   & 96.300\%  & $\sim$1 hour\\
\end{tabular}
\end{center}

Table 4 summarizes the accuracy and runtimes of weighted KNN with $K = 4$ and
perceptron with the polynomial kernel and $d = 10$. KNN is clearly more
accurate, however because perceptron is 8x faster and almost as accurate, there
could be many cases where it is a better choice.

In addition to being faster, perceptron likely has a lower memory
footprint. While KNN must store every sample in the training set, perceptron
only stores mistakes which will be a subset of the entire training set.

@@

\begin{itemize}
\item Expanded cross validation for KNN using the full training set. This would
  help validate that the weighted voting algorithm with $K = 4$ leads to the
  most accurate. 
\item More kernels and a greater assortment of $\sigma$ and $d$ values to
  see if we can increase the effectiveness of the perceptron algorithm. 
\item An SVM implementation which regularizes the data to avoid overfitting. If
  the perceptron implementation overfits the data, this would be a potential
  way to increase the accuracy.
\end{itemize}

>>>>

Optical Character Recognition (OCR) is a complex computer science problem which
has several effective machine learning solutions. We implemented the K Nearest
Neighbors (KNN) and Kernelized Perceptron algorithms and compared the
effectiveness of both methods in classifying a testing data set of 28,000
images. Each image is formatted as an array of 768 pixel values, with each value
indicating how dark each pixel was in the image. Both of our algorithms were
trained on a training data set of 42,000 images. We used cross validation to
tune the K value in KNN and to choose a kernel function and dimension value for
Perceptron. Despite its relative simplicity, we found that KNN was more
effective than Perceptron in classifying the digits. The nearest neighbors
algorithm correctly classified 96.857\% of the testing samples on a weighted
neighbors algorithm which looked at 4 neighbors. Our best Perceptron
implementation correctly classifies 96.3\% of the testing samples using an
exponential kernel with a $\sigma = 10$. Despite its higher accuracy, the runtime
of KNN was 8x slower than that of the Perceptron.

@@

The Kaggle data provides two groups of data, 42,000 training samples with labels
and 28,000 unlabeled testing samples. Each training sample contains a 48 pixel
by 48 pixel image containing a handwritten number. The image is represented as
an array of 768 pixel values which represent how dark each pixel of the image
is. Table 1 is an example of a training data sample.

@@

K Nearest Neighbors is a relatively simple algorithm which often gets
surprisingly good results given its simplicity. Given a testing sample $X$, the
algorithm looks through all the training samples it has previously seen and
finds the \textit{K} most similar samples, known as the nearest neighbors. The
algorithm then implements some form of a voting mechanism among the \textit{K}
samples to classify the testing sample as a digit 0-9. The voting mechanism can
be as straightforward as selecting the majority label, or it can be more
complex, involving weighting samples based on some distance metric.

@@

The weight function is an inverse distance function of the form:

\begin{center}
$Weight(TrainY, TrainX, TestX) = \frac{1}{Euclidean(TestX, TrainX)}*TrainY$
\end{center}

@@

\begin{table}[ht]
\textbf{\caption{KNN Cross Validation Error Rates}} % title
\centering
\begin{tabular}{c | c c }
K & Unweighted KNN Error & Weighted KNN Error \\
\hline
1 & 7.7\% & 7.7\% \\
2 & 9.2\% & 9.2\% \\
3 & 7.5\% & 7.5\% \\
4 & 7.8\% & 7.3\% \\
5 & 7.8\% & 7.5\% \\
10 & 8.5\% & 7.9\% \\
15 & 9.5\% & 8.9\% \\
20 & 10.2\% & 9.8\% \\ [1ex]
\end{tabular}
\label{table:nonlin}
\end{table}

@@

Our cross validation results suggest that running weighted KNN with $K = 4$
yields the most successful classification rate. One caveat to this is that our
cross validation code ran using 5,000 training samples instead of the 42,000
samples in our full training set. It is possible that data we used to cross
validate is not representative of the full data set.

@@

Our KNN implementation was extremely successful, correctly classifying 96.857\%
of the testing samples. One downside to it, however, is the runtime. Training is
near instantaneous since the algorithm simply loads the data into an array. To
classify a sample the algorithm needs to compute the euclidean distance between
the testing sample and every training sample. In the case of the full data set
this is 42,000 samples, each of which has 768 dimensions. As a result, each
testing sample takes approximately one second to run. Altogether it takes close
to 8 hours to classify all 28,000 testing samples.

@@

One problem that needs to be addressed is that a standard perceptron is a binary
classifier meaning it selects between one of two classes. We resolved this issue
by training $10 \choose 2$ $=45$ classifiers. Each classifier is responsible for
classifying between two digits. For example, we would train a perceptron to
classify between 0 and 1, and a second to classify between 0 and 2, and so
on. When we classify we vote between all of our trained classifiers and output
the majority element.

@@

The kernel function is represented by $K(X_i, X)$. There are a number of choices
for this kernel function which we will explore in the next section.

@@

Finally, the runtime of the perceptron algorithm is a little over an hour. It
takes approximately 45 minutes to train and about 20 minutes to test the
data. The faster runtime allowed us to cross validate on the full data set so
our cross validation results are more accurate than those we obtained for KNN.

@@

\begin{table}[ht]
\textbf{\caption{KNN vs. Perceptron on Full Testing Set}}
\centering
\begin{tabular}{ c| c c}
  \hline
  Algorithm & Percent Success & runtime\\
  \hline
  Nearest Neighbors, $k = 4$   & 96.857\% & $\sim$8 hours\\
  Perceptron   & 96.300\%  & $\sim$1 hour\\
\end{tabular}
\end{table}

Table 4 summarizes the accuracy and runtimes of weighted KNN with $K = 4$ and
perceptron with the polynomial kernel and $d = 10$. KNN is clearly more
accurate, however because perceptron is nearly 8x faster, there could be many
cases where it is a better choice.

In addition to being faster, perceptron likely has a lower memory
footprint. While KNN must store every sample in the training set, perceptron
only stores mistakes which will be a subset of the training set.

One reason why KNN may have been so successful in this trial is due to the large
training data set and small number of classes to classify. With only 10 posible
classifications, there are likely around 4,2000 training samples of each
number. The odds of at least a few of those being similar to the test image
intuitively seem high.

An explanation for the perceptron algorithm being less successful than KNN is
that it may have overfit our training data because there was no
regularization. This could explain why our polynomial kernel did worse with a
higher dimension kernel function.

@@

\begin{itemize}
\item Expanded cross validation for KNN using the full training set. This would
  help validate that the weighted voting algorithm with $K = 4$ leads to the
  most accurate classification. 
\item More kernels and a greater assortment of $\sigma$ and $d$ values to
  see if we can increase the effectiveness of the perceptron algorithm. 
\item An SVM implementation which regularizes the data to avoid overfitting. If
  the perceptron implementation overfits the data, this would be a potential
  way to increase the accuracy.
\end{itemize}
